---
title: "Beginner Tutorial"
author: "Livia Popa & Nick Gawron"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(rnoaa)
# uploads library info
```

# Beginner Tutorial

# Open Data Science

## What is Open Data and Reproducible Science?

Open data documents and shares research data openly for re-use.

Open data research aims to transform research by pushing change in the way that research is carried out and disseminated by digital tools. Open data should be:

- Publicly available: Open data is freely available on the internet.
- Reusable: Proper licensing is essential for research outputs so that users know any limitations on re-use
- Transparent: With appropriate metadata to explain how research output was produced and what it contains


## Importing Data From Cardinals

- Cardinal is a high-powered, user-oriented, one-stop-shop for North Carolina weather and climate data housed at the North Carolina State Climate Office. 
- Cardinal makes weather and climate data more accessible to users, with features and prompts that take the guesswork out of station and parameter identification and selection.

## Using data inspection functions

### Data Types

- Date
- Str
- Char
- Double 
- Num

```{r}
library(readr)
cardinal <- read_csv("cardinal_data.csv")
#cardinal%>%filter(colnames(cardinal)!="MV")
```


```{r echo=F}
cardinal_data <- read_csv("cardinal_data.csv", 
     col_types = list(`Average Air Temperature (F)` = col_number(), 
         `Maximum Air Temperature (F)` = col_number(), 
         `Minimum Air Temperature (F)` = col_number(), 
         `Average Experimental Leaf Wetness (mV)` = col_number(), 
         `Total Precipitation (in)` = col_number(), 
         `Average Relative Humidity (%)` = col_number(), 
         `Average Soil Moisture (m3/m3)` = col_number(), 
         `Average Soil Temperature (F)` = col_number(), 
         `Average Solar Radiation (W/m2)` = col_number(), 
         `Average Station Pressure (mb)` = col_number()))

cardinal_data<-drop_na(cardinal_data)
str(cardinal_data)
cardinal_data$Date<-as.Date(cardinal_data$Date, tryFormats= c("%m/%d/%y"))
view(cardinal_data)

#changes col names
colnames(cardinal_data)=c("date","AvgT","MaxT","MinT","AvgLw","Tprep","AvgHum","AvgSm","AvgSt","AvgSr","AvgStp")
```


## Making new Data

```{r}
#me making new data

cardinal_data$IfRain<- (cardinal_data$Tprep>0)
cardinal_data$IfRain<-as.factor(as.integer(cardinal_data$IfRain))
cardinal_data 

```




## Numerical Summaries





```{r}
# base plots in R, categorical variables
#does a count
plot(cardinal_$IfRain)
```

## Plotting Basic 

<<<<<<< Updated upstream
We will be using a package called ggplot2.
=======

We will be using a fancy package called ggplot2.
>>>>>>> Stashed changes

Here is a good link: https://www.rstudio.com/resources/cheatsheets/

```{r}
ggplot(cardinal_data,aes(x= date ,y= AvgLw))+ geom_point()
```


```{r}
#livia
ggplot(cardinal_data,aes(x=AvgT)) + geom_histogram(color="green",fill="orange")+labs(x="Average Temperature", y="Day",title ="Histogram of Average Temperature in Lake Wheeler")

ggplot(cardinal_data,aes(x=date,y=AvgT,fill=IfRain))+geom_area()
```


```{r}
#me 
ggplot(cardinal_data,aes(x=date,y=AvgT))+geom_line(aes(col=IfRain))
```




